Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-27 sections/cut.tex
formatting

>>>>

%
\teo{
The Finch language offers extensible abstractions for elements of control flow (conditionals and indexing expressions) that the Finch compiler predictably converts into the same structured representation used to describe arrays.
%
By controlling how control flow is lowered into this representation, a programmer can control how structured data is combined with control flow to produce a single control flow, the program.
%
We observe that this strategy naturally extends the dense solution.
}
\teo{I think we can back it up in two ways:
1. We need to modify the semantics to focus when on when things are converted into Looplets
2. We need to highlight the specifics in the case studies: in particular, the blur and the symmetric operations. If we do triangle counting or something like that, we could maybe talk about that a bit.
}
\teo{
\item Extensibility of the language and compiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
}

### 2024-03-27 sections/intro.tex
Update on Overleaf.
formatting

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
% EASY CUT
These frameworks have been the
subject of extensive industry investment to enable performant implementations,
and often operate at the peak capacity of the hardware they run on
\cite{lo_roofline_2015}.

@@

\begin{table}[h!]
  \centering
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
  \end{table}

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}

@@

As an example, a programmer might explore the different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices). The control flow here is only useful if the first example differs from the second in that it actually selects only even indices as the two integer lists are merged:
%The Finch compiler uses the structure of the data to generate efficient implementations of these programs.

@@

\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. 
%
To our knowledge, the Finch programming language is the first 
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands. 
%
\item A compiler that specializes programs to data structures 
predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures.
%
%
\teo{
The Finch language offers extensible abstractions for elements of control flow (conditionals and indexing expressions) that the Finch compiler predictably converts into the same structured representation used to describe arrays.
%
By controlling how control flow is lowered into this representation, a programmer can control how structured data is combined with control flow to produce a single control flow, the program.
%
We observe that this strategy naturally extends the dense solution.
}
\teo{I think we can back it up in two ways:
1. We need to modify the semantics to focus when on when things are converted into Looplets
2. We need to highlight the specifics in the case studies: in particular, the blur and the symmetric operations. If we do triangle counting or something like that, we could maybe talk about that a bit.
}
\teo{
\item Extensibility of the language and compiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
}

\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface. 
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

>>>>

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
% EASY CUT
%These frameworks have been the
%subject of extensive industry investment to enable performant implementations,
%and often operate at the peak capacity of the hardware they run on
%\cite{lo_roofline_2015}.

@@

\newcommand*\rot{\rotatebox{90}}

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\end{table}

@@

As an example, a programmer might explore different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices). The control flow here is only useful if the first example differs from the second in that it actually selects only even indices as the two integer lists are merged:
%The Finch compiler uses the structure of the data to generate efficient implementations of these programs.

@@

\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. 
%
To our knowledge, the Finch programming language is the first 
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands. 
%
\item A compiler that specializes programs to data structures automatically,
facilitating an expressive language that makes it easier to search the complex
space of algorithms and data structures. 
%
\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface. 
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

### 2024-03-27 sections/tensors.tex
formatting

\section{Bridging Looplets and Finch: The Tensor Interface}

%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible.
%
First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is best suited towards iterators over a single dimension.
%
We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
%
Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
%
So we require a bridge that can provide different iteration orders depending on the context.
%
Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.

@@

%What needs to be said about the tensors? It's basically a few main points:
%1. (DONE) What is the level abstraction 
%2. (Done) We have identified 8 key level structures to represent most combinations of runs or pinpoints.
%2. a. beautiful figure with datas as rows of larger structures
%2. b. How do we write to each of these, in order? Randomly? random access? several important datastructures that show up along the way.
%3. What is "Unfurling"? and how does it help us represent each structure? refer to looplet decompositions of each case in the earlier figure
%4. What are lifecycles? How do lifecycles help us keep sane? Why are they necessary for correctness?

\subsection{Level Abstraction}
Fiber-tree style tensor abstractions have been the subject of extensive study
\cite{sze2017efficient, chou2022compilation, chou2018format}.  The underlying
idea is to represent a multi-dimensional tensor as a nested vector
datastructure, where each level of the nesting corresponds to a dimension of the
tensor. Thus, a matrix would be represented as a vector of vectors. This kind of
abstraction lends itself to representing sparse tensors if we vary the type of
vector used at each level in a tree. Thus, a sparse matrix might be represented
as a dense vector of sparse vectors. The vector of subtensors in this
abstraction is referred to as a \textbf{fiber}. Prior fiber-tree representations
focus on sparsity (where only the nonzero elements are represented) and treat
sparse vectors as sets of represented points. Since our fiber-tree
represesentation must handle other kinds of structure, such as diagonal,
repeated, or constant values, we instead view each fiber as a mapping from
indices into a space of subfibers.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.45\linewidth]{LevelsVsFibers-matrix.png}\hfill%
    \includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
    \caption{Levels and fiber tree representations of a sparse matrix and a sparse tensor. On left, a matrix is represented in a fibertree corresponding to CSC format, with a dense outer level and a sparse inner level. On right, a tensor is represented in a fibertree with two sparse outer levels, and a dense inner level. Note that the element levels in this case form the leaves of the tree.}
    \label{fig:levelsvsfibers}
\end{figure}

\subsection{The 8 Key Level Structures}
    \begin{wrapfigure}{r}{.35\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{wrapfigure}
    The main benefits of specializing to structure come from the following properties of the data:
    \begin{enumerate}
        \item[Sparsity] Sparse data is data that is mostly zero, or some other
        fill value. When we specialize on this data, we can use annihilation ($x
        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
        redundant work.

>>>>

\section{Bridging Looplets and Finch: The Tensor Interface}
%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible.
%
First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is best suited towards iterators over a single dimension.
%
We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
%
Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
%
So we require a bridge that can provide different iteration orders depending on the context.
%
Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.

\begin{wrapfigure}{R}{0.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
    \label{fig:levelsvsfibers}
\end{wrapfigure}

@@

%What needs to be said about the tensors? It's basically a few main points:
%1. (DONE) What is the level abstraction 
%2. (Done) We have identified 8 key level structures to represent most combinations of runs or pinpoints.
%2. (Done) a. beautiful figure with datas as rows of larger structures
%2. b. How do we write to each of these, in order? Randomly? random access? several important datastructures that show up along the way.
%3. What is "Unfurling"? and how does it help us represent each structure? refer to looplet decompositions of each case in the earlier figure
%4. (Done) What are lifecycles? How do lifecycles help us keep sane? Why are they necessary for correctness?

\subsection{Level Abstraction}

Fiber-tree style tensor abstractions have been the subject of extensive study
\cite{sze2017efficient, chou2022compilation, chou2018format}.  The underlying
idea is to represent a multi-dimensional tensor as a nested vector
datastructure, where each level of the nesting corresponds to a dimension of the
tensor. Thus, a matrix would be represented as a vector of vectors. This kind of
abstraction lends itself to representing sparse tensors if we vary the type of
vector used at each level in a tree. Thus, a sparse matrix might be represented
as a dense vector of sparse vectors. The vector of subtensors in this
abstraction is referred to as a \textbf{fiber}. Prior fiber-tree representations
focus on sparsity (where only the nonzero elements are represented) and treat
sparse vectors as sets of represented points. Since our fiber-tree
represesentation must handle other kinds of structure, such as diagonal,
repeated, or constant values, we instead view each fiber as a mapping from
indices into a space of subfibers.

@@

\subsection{The 8 Key Level Structures}
    \begin{wrapfigure}{R}{.35\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{wrapfigure}
    The main benefits of specializing to structure come from the following properties of the data:
    \begin{enumerate}
        \item[Sparsity] Sparse data is data that is mostly zero, or some other
        fill value. When we specialize on this data, we can use annihilation ($x
        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
        redundant work.
